### 2021-12-14 text/jsm/jsm_paper.tex
update appendix

\documentclass[11pt]{article}
\usepackage{setspace}
\doublespacing
\usepackage[margin=1in]{geometry}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{array}

\usepackage[
backend=biber,
style=numeric-comp,
maxcitenames=1,
maxbibnames=3,
backref=true,
doi=false,
isbn=false,
url=false,
eprint=false
]{biblatex}

\addbibresource{/Users/timbarry/optionFiles/glmeiv.bib}
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
\usepackage[toc,page]{appendix}
\usepackage{algpseudocode}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\makeatletter
\newcommand{\multiline}[1]{%
	\begin{tabularx}{\dimexpr\linewidth-\ALG@thistlm}[t]{@{}X@{}}
		#1
	\end{tabularx}
}
\makeatother
\setcounter{tocdepth}{2}
\usepackage{xcolor}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\usepackage{float}
\allowdisplaybreaks
\usepackage[caption = false]{subfig}
\usepackage{/Users/timbarry/optionFiles/mymacros}
\DeclareMathOperator*{\argmin}{arg\,min}

\begin{document}
\noindent
\begin{center} 
\textbf{A new class of exponential family measurement error models applied to single-cell CRISPR screens} \\
Timothy Barry, Kathryn Roeder, Eugene Katsevich
\end{center}

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression in challenging problem settings. We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}

@@

Our work makes two main contributions. First, we conduct a detailed study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting. Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. % Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.

@@

\subsection{Related work}
Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin et al.\ \cite{Lin2021} and Townes et al.\ \cite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions. Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \cite{Lause2021}, improving performance on dimension reduction tasks. We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction. We extend the classical errors-in-variables model in several directions, most importantly generalizing to response distributions and sources of measurement error that are exponential family-distributed.

@@

The genomics literature has produced several applied methods for single-cell CRISPR screen analysis. In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we aim to account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding. Another difference is that we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is more challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.

@@

\section{Thresholding method}
We study thresholding from empirical and theoretical perspectives, highlighting several inherent difficulties of the approach. Gasperini and Xie imputed perturbation assignments onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini fitted negative binomial regression models to the data, whereas Xie applied nonparametric tests of independence. We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV and (in our view) is the more natural method.

@@

\item[3.]  Fit a GLM to obtain an estimate and confidence interval for the target of inference $\beta^m_1$.
\end{itemize}
The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression. Exponentiating both sides of (\ref{thresh_glm}) yields
$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m,$$ where $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.

Fold change in this context is defined as the ratio of the mean expression of perturbed cells to the mean expression of unperturbed cells. The target of inference $\beta^m_1$ is the \textit{log} fold change in expression in response to the perturbation, controlling for the technical factors. Therefore, $\exp(\beta^m_1) = 1$ (eq., $\beta^m_1 = 0$) indicates no change in expression, while $\exp(\beta^m_1) > 1$ (eq., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (eq., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.

\subsection{Empirical challenges of thresholding method}\label{sec:thresholding_empirical}

@@

To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method on the positive control data using three different choices for the threshold: 1, 5, and 20. We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b). Estimates for log fold change produced by threshold = 1 were smaller in magnitude than those produced by threshold = 5. (Equivalently, estimates for \textit{raw} fold change were closer to the baseline of $1$ for threshold = 1; Figure \ref{thresholding_empirical}a.) Estimates produced by threshold = 5 and threshold = 20 were more concordant, but threshold = 20 yielded slightly larger effect sizes (Figure \ref{thresholding_empirical}b).

We reasoned that the thresholding method systematically underestimates effect sizes on the positive control pairs, especially when the threshold is small. For a given perturbation, the vast majority ($>98\%$) of cells are unperturbed. This imbalance leads to an asymmetry: misclassifying \textit{unperturbed} cells as \textit{perturbed} is intuitively ``worse'' than misclassifying \textit{perturbed} cells as \textit{unperturbed}. Misclassified unperturbed cells contaminate the set of truly perturbed cells, leading to attenuation bias; by contrast, misclassified perturbed cells are swamped in number and ``neutralized'' by the truly unperturbed cells. Setting the threshold to a large number reduces the unperturbed-to-perturbed misclassification rate, decreasing bias.

We hypothesized, however, that the reduction in bias obtained by selecting a large threshold comes at the cost of increasing the variance of the estimator. To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 versus threshold = 20 for the target of inference $\beta^m_1$. We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figure \ref{thresholding_empirical}c-d). We concluded that the threshold controls a bias-variance tradeoff: as the threshold increases, bias of the estimator decreases and variance increases.

@@

\subsection{Theoretical challenges of thresholding method}\label{sec:thresholding_theory}

Next, we study the thresholding method from a theoretical perspective, recovering in precise mathematical terms the attenuation bias and bias-variance tradeoff effects uncovered on real data, as well as several other interesting phenomena.  We work in an idealized Gaussian setting. Suppose that we observe gRNA and gene expression data $\{(g_1, m_1), \dots, (g_n, m_n)\}$ on $n \in \N$ cells from the following model:

\begin{equation}\label{theoretical_model}
\begin{cases}
m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i \\
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i \\
p_i \sim \textrm{Bern}(\pi) \\
\ep_i, \tau_i \sim N(0,1) \\
p_i \indep \tau_i \indep \ep_i.
\end{cases}
\end{equation}
For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is given by $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2 }.$$

\begin{proposition}\label{prop:convergence}
 The almost sure limit (as $n \to \infty$) of $\hat{\beta}^m_1$ is
\begin{equation}\label{thresh_est_intercepts}
\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),
\end{equation}
where
$$\begin{cases}
\E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi , \\
\omega = \Phi\left(\beta_1^g + \beta_0^g -c \right) ,\\ \zeta = \Phi\left( \beta^g_0 - c \right). \\
\end{cases}$$
\end{proposition}
Let $\gamma: \R^4 \to \R$ be defined by
$$ \gamma(\beta^g_1, \pi, c, \beta^g_0) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the ``attenuation function.'' Observe that \begin{itemize}
\item[i.] $\gamma$ does not depend on $\beta^m_1$ or $\beta^m_0$, and
\item[ii.] $\hat{\beta}^m_1 \xrightarrow{a.s.} [\gamma(\beta_0^g, \beta_1^g, c, \pi)] \beta^m_1.$
\end{itemize}
Let $b: \R^4 \to \R$ be the asymptotic relative bias of $\hat{\beta}^m_1$:
\begin{multline*}
b(\beta^g_1, \pi, c, \beta^g_0) = \left(\frac{1}{\beta^m_1}\right) \lim_{n\to\infty} \left(\beta^m_1 - \E[\hat{\beta}^m_1]\right) = \left( \frac{1}{\beta^m_1}\right)\left(\beta^m_1 - \E \left(\lim_{a.s.} \hat{\beta}^m_1 \right)\right) \\ = \frac{1}{\beta^m_1} \left(\beta^m_1 - \gamma(\beta^g_1, \pi, c, \beta^g_0)\beta^m_1\right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0),
\end{multline*}
where $\lim_{a.s.}$ denotes a.s.\ convergence. The asymptotic relative bias vanishes when the attenuation function equals $1$.

\begin{center}
\textbf{Bias as a function of threshold (Panel a)}
\end{center}

To investigate the basic question of ``What is a good threshold selection strategy?'', we study the relationship between the asymptotic relative bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$. For simplicity, we begin by setting the perturbation probability $\pi$ to $1/2$. Let $c_\textrm{bayes} \in \R$ be the Bayes-optimal decision boundary for classifying cells as perturbed or unperturbed, i.e. $$c_\textrm{bayes} = \argmin_{c \in \R} \P(\hat{p}_i \neq p_i).$$ Simple algebra shows that $c_\textrm{bayes} = \beta_0^g + (1/2) \beta^g_1.$ Below, we give several results for the asymptotic relative bias $b$ of $\hat{\beta}^m_1$. We refer throughout to Figure \ref{thresholding_theoretical}a, which displays plots of asymptotic relative bias versus threshold for different values of $\beta^g_1$. We sometimes refer to ``asymptotic relative bias'' using the shortened term ``bias'' for succinctness. \textcolor{red}{SHOULD WE MOVE PROP 4-6 	TO THE APPENDIX? THESE PROPS ARE LESS IMPORTANT THAN THE OTHERS AND CONTRIBUTE SOMEWHAT LESS TO THE NARRATIVE.}

\begin{itemize}
\item 
\begin{proposition}\label{prop:att_bias} Fix $\pi = 1/2$. For all $(\beta^g_1, c, \beta^g_0) \in \R^3$, the asymptotic relative bias is positive, i.e. 
$$b(\beta^g_1, 1/2, c, \beta^g_0) > 0.$$
\end{proposition}
The thresholding method incurs strict attenuation bias (i.e., it \textit{under}estimates the true effect size) for all choices of the threshold and over all possible values of the model parameters (Figure \ref{thresholding_theoretical}a). Attenuation bias is a common attribute of estimators that ignore measurement in errors-in-variables models \cite{Stefanski2000a}.
\item
\begin{proposition}\label{prop:monotonic} Fix $\pi = 1/2$. The asymptotic relative bias $b$ decreases monotonically in $\beta_1^g$, i.e.
	$$\frac{\partial b}{\partial(\beta^g_1)}\left(\beta^g_1, 1/2, c, \beta^g_0\right) \leq 0.$$
\end{proposition}
This result formalizes the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated. To visualize Proposition (\ref{prop:monotonic}), one can fix a threshold (e.g., $c = 0$) and scan for bias across the panels.	
\item \begin{proposition}\label{prop:bayes_opt}
For $\pi = 1/2$ and given $(\beta^g_1, \beta^g_0) \in \R^2$, the Bayes-optimal decision boundary $c_\textrm{bayes}$ is a critical value of the bias function $b$, i.e.
$$ \frac{\partial b}{\partial c}\left(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0\right) = 0.$$
\end{proposition}
The Bayes-optimal decision boundary is an optimum (or possibly a saddle point) of the asymptotic relative bias function (Figure \ref{thresholding_theoretical}a, vertical blue lines). Interestingly, $c_\textrm{bayes}$ is in some cases a maximizer of the bias (Figure \ref{thresholding_theoretical}a, left) and in other cases a minimizer of the bias (Figure \ref{thresholding_theoretical}a, right).
\item
\begin{proposition}\label{prop:c_limit_half}
Assume without loss of generality that $\beta^g_1 > 0$, and fix $\pi = 1/2$. As the threshold $c$ tends to infinity, the asymptotic relative bias $b$ tends to $1/2$, i.e.
$$\lim_{c \to \infty} b(\beta^g_1, 1/2, c, \beta^g_0) = 1/2.$$
\end{proposition}
In other words, we always can set the threshold to a large number and attain a relative bias of $1/2$ (Figure \ref{thresholding_theoretical}a, all panels). This result establishes an upper bound on the bias of thresholded regression (under optimal threshold selection strategy).

>>>>

% set document class
\documentclass[11pt]{article}

% load packages
\usepackage{setspace}
\usepackage[margin=1in]{geometry}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{array}
\usepackage{titlesec} 
\usepackage{xcolor}
\usepackage{float}
\usepackage[toc,page]{appendix}
\usepackage{algpseudocode}
\usepackage{afterpage}
\usepackage[
backend=biber,
style=numeric-comp,
maxcitenames=1,
maxbibnames=3,
backref=true,
doi=false,
isbn=false,
url=false,
eprint=false
]{biblatex}

% set options
\doublespacing
\addbibresource{/Users/timbarry/optionFiles/glmeiv.bib}
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\makeatletter
\newcommand{\multiline}[1]{%
	\begin{tabularx}{\dimexpr\linewidth-\ALG@thistlm}[t]{@{}X@{}}
		#1
	\end{tabularx}
}
\makeatother
\setcounter{tocdepth}{2}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\allowdisplaybreaks
\usepackage[caption = false]{subfig}
\usepackage{/Users/timbarry/optionFiles/mymacros}
\DeclareMathOperator*{\argmin}{arg\,min}
\titleformat{\subsection}[runin]
{\normalfont\large\bfseries}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[runin]
{\normalfont\normalsize\bfseries}{\thesubsubsection}{1em}{}
\newcommand\blankpage{%
	\null
	\thispagestyle{empty}%
	\addtocounter{page}{-1}%
	\newpage}

\begin{document}
\noindent
\begin{center} 
\textbf{A class of exponential family measurement error models for CRISPR genome engineering and single-cell sequencing} \\
Timothy Barry, Eugene Katsevich, Kathryn Roeder
\end{center}

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. To overcome these limitations, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables. We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, demonstrating improved performance in challenging problem settings.
\end{abstract}

@@

Our work makes two main contributions. First, we conduct a detailed study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in a simplified Gaussian setting. Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. % Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.

@@

\subsection{Related work}
Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin et al.\ \cite{Lin2021} and Townes et al.\ \cite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions. Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \cite{Lause2021} and improving performance on dimension reduction tasks. We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction. We extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.

@@

The genomics literature has produced several applied methods for single-cell CRISPR screen analysis. In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding. Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.

@@

\section{Thresholding method}
We study thresholding from empirical and theoretical perspectives, highlighting several inherent limitations of the approach. Gasperini and Xie imputed perturbation assignments onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini fitted negative binomial regression models to the data, whereas Xie applied nonparametric tests of independence. We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV and (in our view) is the more natural method.

@@

\item[3.]  Fit a GLM to obtain an estimate and confidence interval for the target of inference $\beta^m_1$.
\end{itemize}
The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression. Exponentiating both sides of (\ref{thresh_glm}) yields
$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ We see that $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.

Fold change in this context is defined as the ratio of mean gene expression in perturbed cells to mean gene expression in unperturbed cells. The target of inference $\beta^m_1$ is the \textit{log} fold change in expression in response to the perturbation, controlling for the technical factors. Therefore, $\exp(\beta^m_1) = 1$ (eq., $\beta^m_1 = 0$) indicates no change in expression, whereas $\exp(\beta^m_1) > 1$ (eq., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (eq., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.

\subsection{Empirical challenges of the thresholding method}\label{sec:thresholding_empirical}

@@

To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method to the positive control data using three different choices for the threshold: 1, 5, and 20. We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b). Estimates for fold change produced by threshold = 1 were smaller in magnitude (i.e., closer to the baseline of $1$) than those produced by threshold = 5. (Figure \ref{thresholding_empirical}a.) Estimates produced by threshold = 5 and threshold = 20 were more concordant (Figure \ref{thresholding_empirical}b).

We reasoned that thresholded regression systematically underestimated true effect sizes, especially for small thresholds. For a given perturbation, the majority ($>98\%$) of cells are unperturbed. This imbalance leads to an asymmetry: misclassifying \textit{unperturbed} cells as \textit{perturbed} is intuitively ``worse'' than misclassifying \textit{perturbed} cells as \textit{unperturbed}. Misclassified unperturbed cells contaminate the set of truly perturbed cells, leading to attenuation bias; by contrast, misclassified perturbed cells are swamped in number and ``neutralized'' by the truly unperturbed cells. Setting the threshold to a large number reduces the unperturbed-to-perturbed misclassification rate, decreasing bias.

We hypothesized, however, that the reduction in bias obtained by selecting a large threshold causes the variance of the estimator to increase. To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 and threshold = 20 for the target of inference $\beta^m_1$. We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figures \ref{thresholding_empirical}c-d). We concluded that the threshold controls a bias-variance tradeoff: as the threshold increases, the bias of the estimator decreases and the variance increases.

@@

\subsection{Theoretical challenges of the thresholding method}\label{sec:thresholding_theory}

Next, we studied the thresholding method from a theoretical perspective, recovering in precise mathematical terms phenomena revealed in the empirical analysis (plus several others). We worked in a simplified Gaussian setting. Suppose we observe gRNA expression and gene expression data $(g_1, m_1), \dots, (g_n, m_n)$ on $n$ cells from the following linear model:
\begin{equation}\label{theoretical_model}
m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i; \quad
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i; \quad
p_i \sim \textrm{Bern}(\pi); \quad
\ep_i, \tau_i \sim N(0,1),
\end{equation}
where $p_i, \tau_i,$ and $\ep_i$ are independent. For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e. $\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right]\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}.$ We begin by deriving the limit of $\hat{\beta}^m_1$: 
\begin{proposition}\label{prop:convergence}
	The almost sure limit (as $n \to \infty$) of $\hat{\beta}^m_1$ is
	\begin{equation}\label{thresh_est_intercepts}
	\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right) \equiv \beta^m_1 \gamma(\beta^g_1, \pi, c, \beta^g_0), 
	\end{equation} where 
	$$
	\E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g -c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$
\end{proposition}
The function $\gamma: \R^4 \to \R$ does not depend on the gene expression parameters $\beta^m_1$ or $\beta^m_0$. The asymptotic relative bias $b: \R^4 \to \R$ of $\hat{\beta}^m_1$ is given by
$$b(\beta^g_1, \pi, c, \beta^g_0)  \equiv \frac{1}{\beta^m_1} \left(\beta^m_1 - \lim_\textrm{a.s.} \hat{\beta}^m_1 \right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0).$$
Having derived an exact expression for  the asymptotic relative bias of  $\hat{\beta}^m_1$, we can prove several results about this quantity. For the sake of simplicity, we consider a slightly specialized setting in which we set $\pi$ to to a specific value (see Appendix \ref{sec:appendix_theory} for details; proofs and detailed statements likewise deferred to Appendix \ref{sec:appendix_theory}).

First, the thresholding estimator strictly \textit{underestimates} (in absolute value) the true value of $\beta^m_1$ over all choices of the threshold and over all possible values of the model parameters. This phenomenon, called attenuation bias, is a common attribute of estimators that ignore measurement in errors-in-variables models \cite{Stefanski2000a}. Second, the magnitude of the bias decreases monotonically in $\beta^g_1$, comporting with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated. Third, the Bayes-optimal decision boundary $c_\textrm{bayes} \in \R$ (i.e., the most accurate decision boundary for classifying cells as perturbed or unperturbed) is a critical value of the bias function. Finally, and most subtly, there is no universally applicable rule for selecting a threshold that produces minimal bias: when $\beta^g_1$ is small, setting the threshold to an arbitrarily large number yields smaller bias than setting the threshold to the Bayes decision boundary; when $\beta^g_1$ is large, the reverse is true.

Next, we studied the variance of the thresholding estimator, considering a slightly simpler model for this purpose. Suppose the intercepts in (\ref{theoretical_model}) are fixed at $0$ (i.e., $\beta^m_0$ = $\beta^g_0$ = 0). For notational simplicity write $\beta_m = \beta^m_1$ and $\beta_g = \beta^g_1.$ The thresholding estimator $\hat{\beta}_m$ is the no-intercept OLS solution $\hat{\beta}_m = \left[\sum_{i=1}^n \hat{p}_i m_i \right]\left[\sum_{i=1}^n \hat{p}_i^2 \right]^{-1}.$ We have the following limiting distributional result.
\begin{proposition}\label{prop:bv_decomp}
	The limiting distribution of $\hat{\beta}_m$ is
	$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m \omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right),$$ where $$l = \beta_m \omega \pi/[\zeta(1-\pi) + \omega \pi]; \quad
	\E[\hat{p}_i] = \pi \omega + (1-\pi) \zeta; \quad
	\omega = \Phi(\beta_g - c); \quad
	\zeta = \Phi(-c).$$
\end{proposition}
This proposition yields an asymptotically exact bias-variance decomposition for $\hat{\beta}_m$ (Figure \ref{thresholding_theoretical}): as the threshold tends to infinity, the bias decreases and the variance increases, consistent with empirical observations that we made in Section \ref{sec:thresholding_empirical}. Overall, our empirical and theoretical analyses reveal that thresholded regression poses several difficulties: thresholding incurs strict attenuation bias (that in some settings is large), and selecting a good threshold is challenging. These considerations motivate our core research question: \textit{does modeling the gRNA counts directly, thereby circumventing thresholding, facilitate estimation and inference in single-cell CRISPR screen analysis}?

\section{GLM-EIV}

To answer this question, we introduce GLM-EIV (GLM-based errors-in-variables), a method that models the entire data-generating process.

\subsection{Model}\label{sec:model}
Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})). We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities. To this end, let the latent variable $p_i \in \{0,1\}$ indicate whether cell $i \in \{1, \dots, n\}$ was perturbed. We model the gene expression counts according to
\begin{equation}\label{glmeiv_model_1}
m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i^m); \quad \log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
\end{equation}
where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in \R^{d - 2}$ are unknown constants. The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$. Next, let $d^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size). The model for the gRNA counts is 
\begin{equation}\label{glmeiv_model_2}
g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right); \quad \log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
\end{equation}
where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \R, \beta^g_1 \in \R, \gamma_g \in \R^{d - 2}$ are unknown constants. We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}. Finally, we model the marginal perturbation probability as $
p_i \sim \textrm{Bern}(\pi),$ where $\pi \in (0,1/2]$. 
 Together, (\ref{glmeiv_model_1}), (\ref{glmeiv_model_2}), and the marginal distribution of $p_i$ define the negative binomial GLM-EIV model. The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model. Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors. See Appendix \ref{sec:zero_inf_model} for a parallel methodology that we developed in which we model the gRNA distribution as (in a certain sense) zero-inflated counts.
